---
title: "Class 11: Genome Informatics and High Througput Sequencing"
author: "Aishwarya Ramesh"
format: pdf
---

# Section 1: Proportion of G/G in a Population

Reading csv file

```{r}
mxl <- read.csv('373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv')
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```
Roughly 14.06% are homozygous for asthma associated gene in MXL population. 


Now let's look at a different population.
Looking at GBR population. 

```{r}
gbr <- read.csv('373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv')

```


FInd proportion of GG in GBR

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childbood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further. 
## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. 

How many samples do we have?


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


```{r}
nrow(expr)
```

We have 462 individuals in this data. 

>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

Finding sample sizes
```{r}
table(expr$geno)
```
Finding median expression levels

```{r}
gg_summary <- summary(expr$exp[expr$geno == 'G/G'])
ag_summary <- summary(expr$exp[expr$geno == 'A/G'])
aa_summary <- summary(expr$exp[expr$geno == 'A/A'])
gg_summary
ag_summary
aa_summary

```

The sample size for A/A is 108, the sample size for A/G is 233, the sample size for G/G is 121. The median expression for G/G is 20.074. The median expression for A/G is 25.065. The median expression for A/A is 31.25. 

```{r}
library(ggplot2)

```


> Q14:  Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

Making boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno)+
  geom_boxplot(notch=TRUE)
```

The expression is relatively higher for A/A compared to G/G. From this, we can conclude that having a G/G is associated with having a lower expression of this particular gene ORMDL3, and thus that there will be differential asthma outcomes for A/A as compared to G/G.  