---
title: "Class 8 Mini-Project: Unsupervised Learning Analysis of Human Breast Cancer Cells"
format: html
author: Aishwarya Ramesh
---

## Exploratory data analysis

### Preparing the Data
First, getting the data:

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```

Next, taking a quick look at the data:

```{r}
head(wisc.df)
```

Excluding the `wisc.df$diagnosis` column in our analysis: 

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

Checking whether it removed: 

```{r}
head(wisc.data)
```

Creating `diagnosis` vector for future comparison:

```{r}
# Create diagnosis vector for later 
diagnosis <- factor(wisc.df[,1])
diagnosis
```

### Exploratory data analysis

> Q1. How many observations are in this dataset?

Finding dimensions of data: 

```{r}
dim(wisc.data)
```
This dataset has 569 observations. 

> Q2. How many of the observations have a malignant diagnosis? 

```{r}
table(diagnosis)
```
There are 212 observations with a malignant diagnosis. 

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
wisc.columns <- colnames(wisc.data)
suffixed_list <- grep('_mean', wisc.columns)
length(suffixed_list)
```


## Principal component analysis

## Performing PCA

Checking whether data needs to be scaled:

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)

```

There seems to be significantly different variances in the variance, so we must scale the data before PCA.

```{r}
scaled_wisc <- scale(wisc.data, center=TRUE, scale=TRUE)
colMeans(scaled_wisc)
apply(scaled_wisc,2,sd)
```

The scaling seems to have made the variances of data more similar. Now we can perform PCA. 

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale=TRUE, center=TRUE)
```

To get summary of principal components, we use `summary()` and we can also use `plot()` to get visual representation of proportion of variance described by each PC. 

```{r}
summary(wisc.pr)
plot(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

From the data above, we see that PC1 captures 44.27% of the original variance in the dataset.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Three principal components (PC1, PC2 and PC3) are required to describe 72.64% of the variance in the originan