---
title: "Class 05: Data Visualization with GGPLOT"
author: "Aishwarya"
format: pdf
---

# Our first ggplot

To use the ggplot2 package I first need to have it installed on my computer.

To install packages we use the `install.packages()` command.

Now can I use it? NO! First need to call `library(ggplot2)`

```{r}
library(ggplot2)
ggplot()
```

```{r}
mpg
```

Our first plot of displ vs hwy All ggplot() graphs are made in the same way:

-   data + aes + geoms

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point()
```


I can add more layers

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE)
```

## Plot of gene expression data

First read the data from online. 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```

What are the colnames?
```{r}
colnames(genes)
```
Creating a scatterplot of Condition1 and Condition2:
```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2) +
  geom_point() #+
  #geom_smooth(method=lm, se=FALSE)
```

Adding color now:
```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
p
```

> Q: How many genes are up regulated and down regulated?

```{r}
head(genes)
```

To get at just the State column
```{r}
table(genes$State)

```

Another plot
```{r}
q <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point(alpha=0.3)
q
```

Making plot more colorful:
```{r}
p <- p + scale_colour_manual( values=c("blue","gray","red") ) + labs(title='Gene Expression Changes Upon Drug Treatment',                              x='Control(No Drug)', y='Drug Treatment', )
p
```


