---
title: "Class 10: Halloween Mini-Project"
author: Aishwarya Ramesh
format: pdf
---

```{r}
# webshot::install_phantomjs()
```

## 1. Importing candy data

```{r}
candy_file <- 'candy_data.csv'
candy <- read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv', row.names=1)
head(candy)
```

> Q1: How many different candy types are in this dataset? 

```{r}
dim(candy)
```

There are 85 different types of candy in this dataset. 

> Q2. How many fruity candy types are in the dataset? 

```{r}
sum(candy$fruity)
```

There are 38 types of fruity candy. 

## 2. What is your favorite candy?

Finding `winpercent` value for Twix. 

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Air Heads. 

```{r}
candy['Air Heads', ]$winpercent
```
The winpercent value is 52.34146 percent. 


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy['Kit Kat', ]$winpercent
```

The winpercent value for Kit Kat is 76.7686 percent. 


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy['Tootsie Roll Snack Bars', ]$winpercent
```

It is 49.6535 percent. 

(Installed skimr)

```{r}
library("skimr")
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The `winpercent` variable is on a different scale, since all other variables measure as a proportion out of 1 or as a binary 0 or 1, while `winpercent` measures as a percent out of 100. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A 0 means that particular candy is not chocolate, while a 1 means that particular candy contains chocolate. 


Making histograms

> Q8. Plot a histogram of winpercent values 


```{r}
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins=10, col='red')
  
```


> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values is not symmetrical. There are more candies that have lower scores than ones that have higher scores. 


> Q10. Is the center of the distribution above or below 50%?

Yes, the center of the distribution is from 35-45%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy? 

```{r}
choc_mean <- mean(candy$winpercent[as.logical(candy$chocolate)])
fruity_mean <- mean(candy$winpercent[as.logical(candy$fruity)])
choc_mean
fruity_mean
```

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

The difference in between the winpercents is statistically significant with p = 2.871e-08 < 0. 


## 3. Overall Candy Rankings

Sorting the whole dataset

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
least_liked <- rownames(candy %>% arrange(winpercent) %>% head(5))
least_liked
```

The five least liked candy types are 'Nik L Nip', 'Boston Baked Beans', 'Chiclets', 'Super Bubble', 'Jawbusters'. 


> Q14. What are the top 5 all time favorite candy types out of this set? 

```{r}
most_liked <- rownames(candy %>% arrange(desc(winpercent)) %>% head(5))
most_liked
```

The top 5 all time favorite candy types are Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat and Snickers. 


Making plots. 

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent? 

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()

```


### Time to add a useful color


Making black bars for every candy and overrwriting by type
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```



```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Filling columns with color

> Q17. What is the worst ranked chocolate candy? 

The worst ranked chocolate candy is Sixlets. 

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 


## 4. Taking a look at pricepoint

Making a plot of winpercent vs pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck? 

Reeses Miniatures seems to be the highest ranked in terms of winpercent for the least money. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip is the least popular of the top 5 most expensive candy types. 


> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Making barplot of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
  

```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()

```


## 5. Exploring correlation structure


```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)? 

`chocolate` and `fruity` are anti-correlated. 

> Q23. Similarly, what two variables are most positively correlated?

`chocolate` and `bar` is an example of two variables that are positively correlated. 


## 6. Principal Component Analysis

Doing PCA on candy dataset. 


```{r}
# if we scale
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Plotting PC1 vs PC2

```{r}
plot(pca$x[,1:2])
```


Coloring by whether its chocolate, fruity etc. 

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

Trying with ggplot2 by making new data-frame

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

Then plotting 

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


Labeling plot with non-overlapping candy names

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

Getting plotly

```{r}
library(plotly)
```

Putting our PCA plot into plotly

```{r}
ggplotly(p)
```


Taking a final look at our PCA variables

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The fruity, hard and pluribus variables are picked up strongly by PC1 in the positive direction. This makes sense because these three variables tend to be correlated. That is, fruity candies tend to be hard or plural, as opposed to a bar. 

